---
title: "Shooting Incident Trends in New York"
author: "RB"
date: "2025-06-16"
output: html_document
---
```{r set_up, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  results = 'markup',
  message = FALSE,
  warning = FALSE 
)
```
## Dataset overview

This analysis examines the data from **NYPD Shooting Incident Data (Historic)**, which is provided by the New York City Police Department (NYPD). You can access the data from [Data.gov](https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic). The analysis will focus on incident patterns regarding victim demographics and temporal patterns.

## Key Concepts
  1. Importing and cleaning data
  2. Demographic analysis - identifying victim age and gender distributions
  3. Temporal patterns - identifying day-of-week effects and forecasting trends.
  4. Statistical testing - Using ANOVA to confirm statistical significance.

## Data Overview
The Data from **NYPD Shooting Incident Data (Historic)** includes incident timing, location, and demographic information for victims and perpetrators.
  
### Research Questions
1. What are the demographics of victims
  - By gender
  - By age group
2. Do incidents occurring follow a pattern based on day of the week or month?

## Step 1: Importing the raw data. 
This data is hosted as a csv file on NYC open data

```{r load_and_import, results='hide'}
#libraries needed
library(tidyverse)
library(janitor)
library(lubridate)
library(ggplot2)

#First, lets import the data
url = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
shooting_data = read_csv(url)
```
```{r glimpse_raw_data}
# Preview the data
head(shooting_data)
```

## Step 2:  **Cleaning** the data (making it tidy)
We'll standardize column names, convert dates to proper formatting, and select relevant variables.

```{r Cleaning}

# column names to all lower case
cleaned_data = shooting_data %>%
      clean_names()

# convert date from string to data using lubridate
cleaned_data <- cleaned_data %>%
  mutate(occur_date = mdy(occur_date))

# select columns of interest only 
selected_data <- cleaned_data %>%
  select(
    occur_date,
    boro,
    vic_age_group,
    vic_sex,
    vic_race,
    perp_age_group,
    perp_sex,
    perp_race,
    statistical_murder_flag
  )

# add time and day columns
selected_data <- selected_data %>%
  mutate(
    year = year(occur_date),
    day_of_week = wday(occur_date, label = TRUE)
  )

# preview data
head(selected_data) 
```

## Step 3: Victim Demoghraphics
  
### Data Quality Check 
**Checking** data for duplicates, unique , or missing information.
```{r checking_data}
# select columns of interest
victim_demo <- cleaned_data %>%
  select(vic_age_group, vic_sex)

# unique characters in age & gender
unique(victim_demo$vic_age_group)
unique(victim_demo$vic_sex)
```

### Distribution of Categories
lets **check on the counts** of each of those found unique charcaters. 
```{r check_counts}
# Count full distribution of age groups
cleaned_data %>%
  count(vic_age_group) %>%
  arrange(desc(n))

# Count full distribution of sex
cleaned_data %>%
  count(vic_sex) %>%
  arrange(desc(n))
```
### Clean Demographic data  
**remove the unknown** data:  As shown above, there are multiple unknowns and 1 typo.
```{r remove_unknown_from_age_gender}
# Filter and clean victim demographic data for plotting
victim_filtered_data <- cleaned_data %>%
  filter(
    vic_age_group %in% c("<18", "18-24", "25-44", "45-64", "65+"),
    vic_sex %in% c("M", "F")
  ) %>%
  select(vic_age_group, vic_sex)
# Check your work
head(victim_filtered_data)
```

### Age Group Distribution
**Bar graph** to show the distribution of victim age groups.
  
```{r bar_plot_for_victim_age}

ggplot(victim_filtered_data, aes(x = vic_age_group)) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.3, size = 4) +
  labs(
    title = "Victim Age Group Distribution",
    x = "Age Group",
    y = "Count"
  ) +
  theme_minimal()
```

### Gender Distribution
**Pie chart** for victim gender 
```{r pie_chart_for_victim_gender}
# Summarize counts and compute percentages
gender_counts <- victim_filtered_data %>%
  count(vic_sex) %>%
  mutate(percent = round(n / sum(n) * 100, 1),
         label = paste0(percent, "%"))

# Plot
ggplot(gender_counts, aes(x = "", y = n, fill = vic_sex)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5)) +
  labs(
    title = "Victim Gender Distribution",
    fill = "Gender"
  ) +
  theme_void()
```

## Step 4: Time series  Model
  
### Data preparation
We'll create a daily dataset to analyze temporal patterns.
  
```{r filter_and_clean}
# STEP 1: Drop missing dates and select distinct incident per date
incident_counts <- cleaned_data %>%
  filter(!is.na(occur_date)) %>%
  distinct(incident_key, occur_date) %>%  # only unique incidents
  count(occur_date)  # count how many unique incidents per day

# STEP 2: Generate complete date range
min_date <- min(incident_counts$occur_date)
max_date <- max(incident_counts$occur_date)

full_dates <- tibble(occur_date = seq.Date(from = min_date, to = max_date, by = "day"))

# STEP 3: Fill in missing dates with zero incidents
incident_daily <- full_dates %>%
  left_join(incident_counts, by = "occur_date") %>%
  mutate(n = replace_na(n, 0),
         day_of_week = wday(occur_date, label = TRUE))

# Preview the result
head(incident_daily)
```

### Day Of Week Pattern Analysis
  
```{r time_plot}
ggplot(incident_daily, aes(x = day_of_week, y = n)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Shooting Incidents by Day of the Week (Unique Incidents)",
    x = "Day of the Week",
    y = "Number of Unique Incidents"
  ) +
  theme_minimal()
```
The boxplot reveals higher incidence rates on weekends compared to weekdays. Let's test if it is a significant difference.

### Statistical Significance Test
``` {r anova_test_of_signifcants}
# One-way ANOVA: Does day of week affect number of incidents?
anova_result <- aov(n ~ day_of_week, data = incident_daily)
summary(anova_result)
```

The p-value is less than 0.05, we can conclude that the day of the week has a **significant** effect on the probability of an incident occurring. 

### Weekly Forecasting Model
```{r arima_plot}
library(forecast)
library(ggplot2)
library(dplyr)
library(lubridate)
library(zoo)

# Step 1: Aggregate by month
monthly_incidents <- incident_daily %>%
  mutate(month = floor_date(occur_date, "month")) %>%
  group_by(month) %>%
  summarise(incidents = sum(n), .groups = "drop") %>%
  arrange(month)

# Step 2: Create monthly time series object
incident_ts_monthly <- ts(monthly_incidents$incidents,
                          start = c(year(min(monthly_incidents$month)), month(min(monthly_incidents$month))),
                          frequency = 12)  # 12 months per year

# Step 3: Fit ARIMA and forecast
model_monthly <- auto.arima(incident_ts_monthly, seasonal = TRUE)
forecast_monthly <- forecast(model_monthly, h = 6)

# Step 4: Detect peaks using rolling average
monthly_incidents <- monthly_incidents %>%
  mutate(roll_mean = rollmean(incidents, 3, fill = NA, align = "center"),
         is_peak = if_else(
           !is.na(roll_mean) &
             lag(roll_mean, 1) < roll_mean &
             lead(roll_mean, 1) < roll_mean,
           TRUE, FALSE),
         label = if_else(is_peak, format(month, "%b %Y"), NA_character_))

# Step 5: Prepare forecast as data frame
forecast_df <- data.frame(
  month = seq(max(monthly_incidents$month) + months(1), by = "month", length.out = 6),
  forecast = as.numeric(forecast_monthly$mean),
  lower = as.numeric(forecast_monthly$lower[, 2]),
  upper = as.numeric(forecast_monthly$upper[, 2])
)

# Step 6: Plot
ggplot() +
  geom_line(data = monthly_incidents, aes(x = month, y = incidents), color = "black") +
  geom_line(data = forecast_df, aes(x = month, y = forecast), color = "blue") +
  geom_ribbon(data = forecast_df, aes(x = month, ymin = lower, ymax = upper), alpha = 0.2, fill = "blue") +
  geom_point(data = monthly_incidents %>% filter(is_peak), aes(x = month, y = incidents), color = "red", size = 2) +
  geom_text(data = monthly_incidents %>% filter(is_peak),
            aes(x = month, y = incidents, label = label),
            vjust = -0.7, angle = 45, size = 3.2, check_overlap = TRUE) +
  labs(title = "Forecast of Monthly Shooting Incidents in NYC (Peak Months Labeled)",
       x = "Month", y = "Number of Incidents") +
  theme_minimal()
```
The time series model reveals clear long-term trends where incidents are historically peaks during summer months. also shows that incident occurrences are expected to rise for the next couple of weeks if they follow the historical pattern.
  
## Summary
### Key Findings
  - Young Adults (18-44) represent the majority of victims
  - Recorded male victims drastically outnumber female victims
  - Incidents occurred on weekends more than on weekdays.
  - Day of the week patterns are statistically significant.
  - The time series forecasting model shows the continuation of rising historical trends. 
  
### Potential Bias Consideration
  - Data quality issues with unknown/missing demographic information.
  - Reporting bias may exist within certain areas or neighborhoods.
  - Historical data may not consider policy or social changes over the years.
  - The model assumes that past patterns will continue regardless of unaccounted-for future unknown external factors.

This analysis can help create preventive measures to control shooting incidents in NYC.
  
  
  
  
  